---
title: "REPAC quick start guide"
author: 
  - name: Eddie Luidy Imada
    affiliation:
    - &wcm Division of Computational and Systems Pathology, Department of Pathology and Laboratory Medicine, Weill Cornell Medicine.
    email: eimada@med.cornell.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('recount3')`"
vignette: >
  %\VignetteIndexEntry{REPAC quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
The `r Biocpkg('REPAC')` R/Bioconductor package provides an implementation of the method described in Imada et al (2022) for the detection of Differential Polyadenylation Usage (DPU).

# Basics
## Installing REPAC

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then run the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

devtools::install_github('eddieimada/REPAC')
```

## recount3 integration

The `r Biocpkg('REPAC')` package/method is designed to take advantage of the recount3 project by directly extracting expression information for over 750,000 samples.

We recommend the user to get familiar with the project [**recount.bio**](https://LieberInstitute.github.io/recount3-docs) and `r Biocpkg('recount3')` R/Bioconductor package.

# Quick start

After installing `r Biocpkg('REPAC')` you can load the package.

```{r 'start', message=FALSE}
## Load REPAC R package
library("REPAC")
```

REPAC uses the `r Biocpkg('megadepth')` to extract the counts from recount3 bigWig files. Before starting make sure that megadepth is properly installed.

```{r 'install-megadepth', message=FALSE}
## Load REPAC R package
megadepth::install_megadepth()
```

The recount3 project has thousands projects readily available. You can obtain a list of the available projects using the recount3 package.

```{r 'projects'}
## Find all available mouse projects
mouse_projects <- available_projects(organism = "mouse")
```

In order to quantify PAS, REPAC needs a bed file with the positions of small intervals upstream of the PAS. We provide pre-annotated files for hg38 and mm10 genomes (see Imada et al. for details).

```{r 'create_rse'}
## Define the number of parallel cores
future::plan(multisession, workers = 8)

## Select the project you are interested in,
## here we use SRP009615 as an example to
## create a RangedSummarizedExperiment (RSE) object with counts extracted from recount3
rse <- create_pa_rse(organism = "mouse", project="SRP048707", annotation="data/mm10.bed", bed_cols= c("seqnames", "start", "end", "ids", "score", "strand", "gene_name", "type")
)
## Subset groups of interest
rse <- rse[,grepl("WT", rse$sra.sample_title)]

## Filter low expressed PA sites
keep <- edgeR::filterByExpr(se, min.count = 10)
rse <- rse[keep,]

## Test for DPU
results <- fit_corepad(rse, group="sra.sample_title", covariates = NULL)

## Restore plan to single-core
future::plan(sequential)
```


