---
title: "REPAC quick start guide"
author: 
  - name: Eddie Luidy Imada
    affiliation:
    - &wcm Division of Computational and Systems Pathology, Department of Pathology and Laboratory Medicine, Weill Cornell Medicine.
    email: eimada@med.cornell.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('recount3')`"
vignette: >
  %\VignetteIndexEntry{REPAC quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
The `r Biocpkg('REPAC')` R/Bioconductor package provides an implementation of the method described in Imada et al (2022) for the detection of Differential Polyadenylation Usage (DPU).

# Basics
## Installing REPAC

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then run the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

devtools::install_github('eddieimada/REPAC')
```

## recount3 integration

The `r Biocpkg('REPAC')` package/method is designed to take advantage of the recount3 project by directly extracting expression information for over 750,000 samples.

We recommend the user to get familiar with the project [**recount.bio**](https://LieberInstitute.github.io/recount3-docs) and `r Biocpkg('recount3')` R/Bioconductor package.

# Quick start

After installing `r Biocpkg('REPAC')` you can load the package.

```{r 'start', message=FALSE}
## Load REPAC R package
library("REPAC")
```

REPAC uses the `r Biocpkg('megadepth')` to extract the counts from recount3 bigWig files. Before starting make sure that megadepth is properly installed.

```{r 'install-megadepth', message=FALSE}
## Load REPAC R package
megadepth::install_megadepth()
```

The recount3 project has thousands projects readily available. You can obtain a list of the available projects using the recount3 package.

```{r 'projects'}
## Find all available mouse projects
mouse_projects <- available_projects(organism = "mouse")
```

In order to quantify PAS, REPAC requires a `GRanges` (see `r Biocpkg('GenomicRanges')`) object with the positions of intervals upstream of the PAS. We provide pre-annotated 50bp intervals for the hg38 and mm10 genomes (see Imada et al. for details).

```{r 'create_rse'}
## Define the number of parallel cores
future::plan(multisession, workers = 8)
# get mm10 PAS
data(mm10_pa)
## Select the project you are interested in,
## here we use SRP009615 as an example to
## create a RangedSummarizedExperiment (RSE) object with counts extracted from recount3
rse <- create_pa_rse(organism = "mouse", project="SRP048707", annotation=mm10_pa)
```

```{r 'analysis'}
## Subset groups of interest
rse <- rse[,grepl("WT", rse$sra.sample_title)]
## Rename groups
rse$sra.sample_title <- dplyr::case_when(
    grepl("Ex vivo", rse$sra.sample_title) ~ "Control",
    grepl("LPS", rse$sra.sample_title) ~ "LPS")
## Filter low expressed PA sites
keep <- edgeR::filterByExpr(rse, min.count = 10)
rse <- rse[keep,]

## fit model and test for DPU
groups <- rse$sra.sample_title
dMat <- model.matrix(~ 0+groups)
colnames(dMat) <-  gsub("groups", "",colnames(dMat))
cMat <- limma::makeContrasts(
    levels=colnames(dMat),
    LPSvsControl= LPS - Control
)
## Test for DPU
results <- fit_repac(rse, dMat, cMat)
## Resolve how PAS are compared
results <- resolve_comparisons(results)

## Restore plan to single-core
future::plan(sequential)
```

# External datasets

REPAC was originally designed to take advantage of the recount3 resources. Although recount3 collection encompasses over 750,000 samples users may wish to use REPAC with their own data or a data set not included in recount3. This can be done by quantifying the PAS intervals with an external tool, such as `r Biocpkg('Rsubread')`. REPAC provides a wrapper function around the `featureCounts` function to provide an output compatible with `fit_repac`.

```{r 'external-data', eval=FALSE}
se <- rse_from_bam(bams, hg38_pa, pheno, strandSpecific = 0,
                   allowMultiOverlap = F, fraction = F,
                   countMultiMappingReads=F, isPairedEnd = T, 
                   nthreads = 8, countReadPairs = T,  
                   requireBothEndsMapped = F)
````

# Additional information

Q: Can I provide intervals longer than 50bp?
A: Yes. REPAC will extract the counts for any interval provided to it. It is up to the user to make sure these intervals provides a reasonable quantification of polyA activity. We have not tested other interval sizes, so we recommend caution when altering the original strategy.

Q: Does REPAC supports stranded libraries?
A: Yes and no. Due to technical constraints, recount3 data is stored in a bigWig format, which does not support quantification by strand. Nevertheless, the user is still able to perform stranded quantification through other venues (see External datasets) and provide it to the `fit_repac` function.


